### 2019-06-03 2-Chapters/2-Chapter/chapter2.tex
Done for now

We give an overview of the main families of algorithms designed to tackle stochastic MAB problems, and we mainly focus on the \UCB, \klUCB{} and Thompson sampling algorithms, as they are used in the rest of the manuscript.
% We present SMPyBandits in the next Chapter~\ref{chapter:3}, a Python open-source library, designed to run numerical simulations of MAB problems, one of the main contributions of this thesis.
% We use it to compare the empirical performances of the most famous and most efficient algorithms, in terms of regret and by evaluating them on different problems.
% %
% In the point-of-view of numerical simulations or real-world applications of bandits on embedded systems, one must also take care of two other measures of performance: time and memory complexities, which can vary greatly between different families of algorithms.
% We also empirically compare different algorithms for these two measures in Section~\ref{sec:3:timeAndMemoryCosts}.
%
We then give regret upper-bounds of the two main algorithms used in this thesis, \UCB{} and \klUCB, to highlight the difference between their finite-time regret upper-bounds. They both attain a logarithmic regret, proving their order optimality.
% (\ie, their regret upper-bound asymptotically matches the lower-bound up-to a constant factor).
% Moreover, \klUCB{} is known to be optimal for a wide range of arms distributions.
% A sketch of the proof of both results is given in Appendix.

@@

\paragraph{Decisions can use the past observations.}

@@

% Other popular applications include the following.
%
% \begin{itemize}
%     \item
    \textbf{A/B testing}, for instance for websites, is a popular application of the best arm identification problem,
    where the task is purely exploratory and the player is asked to identify the best of two options (or more), in a finite number of steps \cite{audibert2010best}.
    The problem can either consider a fixed budget and no freedom on the ending of the game (\ie, an known horizon $T$), or a fixed confidence and a certain freedom on the budget (\ie, the identified arm must be the true best arm with probability at least $1-\delta$) \cite{Garivier16BAI}.
    The theoretical complexity of this use of bandit for A/B testing was first studied in \cite{Kaufmann14},
    and a more practical point-of-view was proposed in \cite{Jamieson17ABTest}.

@@

\begin{lemma}[Regret decomposition]\label{lem:2:RegretDecomposition}
    The (mean) regret $R_T^{\cA}$ can be decomposed as a sum of the number of selections of sub-optimal arms $k$,
    weighted by their gap, like this:
    \begin{equation}\label{eq:2:RegretDecomposition}
        R_T^{\cA} = \sum_{k=1}^K \Delta_k \; \E[ N_k(T) ]
        = \sum_{\substack{k=1,\dots,K\\\Delta_k > 0}} \Delta_k \; \E[ N_k(T) ].
    \end{equation}
\end{lemma}
%
\begin{proof}\label{proof:2:RegretDecomposition}
    One can use the chain rule of expectation, and because the expectation is taken on the randomness of the (\iid) rewards $(r_k(t))_t$ and on the decisions of the player $(A(t))_t$ (which are measurable wrt to the past observations $\cO_t$),
    one can rewrite the expected cumulated rewards like this,
    $\E \left[ \sum_{t=1}^T r_{A(t)}(t) \right]$
    $= \E \left[ \sum_{k=1}^K \sum_{t=1}^T r_k(t) \mathbbm{1}(A(t) = k) \right]$
    $= \sum_{k=1}^K \E [ r_k(1) \sum_{t=1}^T \mathbbm{1}(A(t) = k) ]$
    % \underbrace{XXX}_{= N_k(T)}
    $= \sum_{k=1}^K \mu_k \E \left[ N_k(T) \right].$
    %
    Thus $R_T^{\cA} = \sum_{k=1}^K \Delta_k \E[ N_k(T) ]$.
    The sum can then be simplified to only account for sub-optimal arms, \ie, arms $k$ such that $\Delta_k > 0$.
\end{proof}

@@

% - Worst case regret (max $R_T^{r}$ for $r$ index of Monte-Carlo simulation?)
The \emph{switching cost} $SC^{\cA}(T)$ counts how many times the player's decision has changed from one time step to the next one, \ie, $SC^{\cA}(T) = \sum_{t=1}^{T-1} \indic(A(t) \neq A(t+1))$.
It has recently gained interest in the literature, for instance in \cite{Koren17}.
% FIXME give the main result proved in this paper?
In single-player models,
it is easy to show that achieving logarithmic regret directly implies a logarithmic upper-bound on the switching cost, and conversely the lower-bound from \cite{LaiRobbins85} also gives an asymptotic logarithmic lower-bound.
But even if $SC^{\cA}(T) = \Theta(\log T)$ for an efficient algorithm, it can be interesting to numerically evaluate this quantity, as a large value might indicate an algorithm that is alternating too much between the optimal arm and other arms.
%
% FIXME we give notations but I don't use them in Chapter5 
It is more interesting to studying $SC_M^{\cA}(T)$, the sum of the switching costs of the $M$ players in a multi-player bandit game, as in this case, it is possible that the players follow an efficient centralized or decentralized algorithm (which gives a logarithmic centralized regret) while still suffering from a linear switching cost.
It is very satisfying to obtain a logarithmic switching cost for our algorithm \MCTopM-\klUCB{} in Chapter~\ref{chapter:5}, even if minimizing the switching cost was not one of our objective, and the bound we obtain on $SC_M^{\cA}(T)$ is just a consequence of our proof of the control we give on the regret $R^{\cA}(T)$.
% - Switching cost? \cite{modiDemo2016} \cite{Koren17}
On a more experimental note, it was showed in \cite{modiDemo2016} that the previous state-of-the-art policy for multi-player bandits, \rhoRand, could be tuned to run in batches\footnote{~The batch bandit setting means that all players use the same decisions for instance for $50$ consecutive times, and update their decisions only once every $50$ time steps. For more details, see \cite{modiDemo2016} for experiments on the multi-player case, or \cite{perchet2016,gao2019batched,kolnogorov2019multi} for theoretical developments on the single-player case.}, in order to reduce by a certain multiplicative factor its switching cost while only adding an additive factor on its regret.

@@

Let us first describe two naive strategies, that both fail dramatically.
We recall the notations introduced above in Section~\ref{par:2:interactiveDemoDiscoverMAB}, the sums of rewards are $X_k(t) = \sum_{s=1}^t r_k(s) \mathbbm{1}(A(s) = k)$, and the numbers of samples are $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$.
%
The estimated means, or empirical averages, are $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$).

% Why Follow-the-Leader (EmpiricalMeans https://smpybandits.github.io/docs/Policies.EmpiricalMeans.html) don't work, example.
-- The ``\emph{Follow-the-Leader}'' strategy consists in first playing once each arm, then always playing $A(t)\in\argmax \widehat{\mu_k}(t)$.
The player only exploits the collected information, and this strategy can fail dramatically, \ie, obtain linear regret in some problems.
Indeed consider $K=2$ Bernoulli arms of means $\mu_1=1/2$ and $\mu_2=\varepsilon$ where $\varepsilon < 1/2$, then with probability $\varepsilon/2$ the player observes a reward of $0$ for arm $1$ then $1$ for arm $2$ on the first rounds, and so she will play arm $2$ for the $T-2$ remaining rounds, giving a linear (mean) regret $R_T \geq \frac{\varepsilon}{2}\left(\frac{1}{2} - \varepsilon\right) (T-1)$.

-- The uniform strategy always plays the $K$ arms uniformly at random.
The player only explores without using the collected information, and this strategy fails dramatically for any non trivial problem (\ie, linear regret).
Indeed it obtains a linear (mean) regret $R_T = \frac{1}{K} \sum_{k=1}^K \Delta_k T$
which gives $R_T \propto T$ for any problem with at least one sub-optimal arms (\ie, all problems expect where $\mu_1=\mu_2=\dots=\mu_K$).

@@

As illustrated by the two previous examples, an efficient strategy needs to solve the trade-off between exploration and exploitation.
The two following solutions both consist in splitting the $T$ time steps into $T_0$ steps of exploration and $T-T_0$ steps of exploitation.
%  either in an alternative way or in a fixed way (``explore then exploit'').

% https://smpybandits.github.io/docs/Policies.EpsilonGreedy.html
-- The $\varepsilon$-greedy strategy consists in alternating exploration and exploitation at a certain ratio \cite{SuttonBarto2018,Bubeck12,LattimoreBanditAlgorithmsBook}.
Fix $0<\varepsilon<1$, then at each round, with probability $\varepsilon$ the player selects an arm uniformly at random (exploration) and with probability $1-\varepsilon$ the arm with highest empirical mean is selected (exploitation).
If $\varepsilon$ is constant, then the (mean) regret is still growing linearly as it is lower bounded by $\varepsilon \frac{1}{K} \sum_{k=1}^K \Delta_k T$.
In this first case, in average $T_0 = \varepsilon T$ steps are spent on exploration.
%
But if we consider $\varepsilon_t$ decreasing with time $t$, for instance $\varepsilon_t = \varepsilon_0 \frac{1}{t}$ with $\varepsilon_0 = 6 K / d^2$, for a constant $0 < d < \min_{k: \Delta_k > 0} \Delta_k$,
then it was shown in \cite{Auer02} that the regret is of the order of $K \log(T) / d + \smallO{T}$, which leads to an order-optimal regret of $\bigO{\log(T)}$.
In this second case, about $T_0 \leq \varepsilon_0 \log(T)$ steps are spent on exploration.

% https://smpybandits.github.io/docs/Policies.ExploreThenCommit.html
-- The ``explore-then-exploit'' strategy, as studied for instance in \cite{GarivierETC2016}, consists in first exploring uniformly the $K$ arms for $T_0$ time steps, then only exploiting the best arm.
We can lower bound the regret of this strategy by
$R_T \geq K (\min_{k: \Delta_k > 0} \Delta_k) T_0/K + p (T - T_0)$,
if $p$ denotes the probability that the chosen arm is not optimal (\ie, the arm with highest mean after $T_0/K$ samples of each arms).
And we can show that $p$ only depends on the number of collected samples $T_0$ and the gaps $\Delta_k$,
so if $T_0$ is fixed independently of $T$ and the problem difficulty, the regret is again growing linearly, \ie, $R_T = \Omega(T)$.

\begin{proof}
    We prove it for the case of two arms.
    If the distributions are Bernoulli (or sub-Gaussian), we can use Hoeffding's inequality from \cite{hoeffding1963probability} to control this probability $p$.
    %
    Indeed if $Z_1,\dots,Z_n$ are \iid{} samples from a Bernoulli distribution of parameter $\mu$, of mean $\widehat{Z_n}$ we have
    \begin{equation}
        \forall x < \mu, \bP(\widehat{Z_n} < x) \leq \exp(-2 n (x-\mu)^2),
        \;\;\;
        \forall x > \mu, \bP(\widehat{Z_n} > x) \leq \exp(-2 n (x-\mu)^2).
    \end{equation}
    %
    If $\mu_1 = \mu_2 + \Delta$, then
    $p = \bP(\widehat{\mu_1} < \widehat{\mu_2}) \leq \bP(\widehat{\mu_1} < \mu_1 - \Delta/2) + \bP(\widehat{\mu_2} > \mu_2 + \Delta/2)$, and both terms can be bounded by using Hoeffding's inequality with $n=T_0/2$ samples,
    to obtain $p \leq 2 \exp(-T_0 \Delta^2 / 4)$.
    %
    For $K$ arms, the same proofs applies, and we can obtain
    $p \leq 2 K \exp(-T_0 d^2 / 4)$, if $d$ is a lower-bound on the positive gaps.
    %
    In all cases, it shows that $p$ is a constant wrt $T$.
\end{proof}

@@

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
            \uIf{$t \leq K$}{
                Play arm $A(t) = t$\;
            }
			\Else(){
                For each arm, compute the index $U_k(t)$\;
                Play arm $A(t) = \arg\max_{1\leq k \leq K} U_k(t)$\;
            }
            Observe a reward $r_{A(t)}(t)$\;
            Update both $X_k(t)$ and $N_k(t)$\;
		}
		\caption{A generic index policy, using indexes $U_k(t)$ (\eg, \UCB, \klUCB{} etc).}
		\label{algo:2:indexPolicy}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

Let us consider another approach for adaptive exploration, known as ``optimism under uncertainty'': assume each arm is as good as it can possibly be given the observations so far, and choose the best arm based on these optimistic estimate.
%
This intuition leads to the $\UCB_1$ algorithm, initially introduced in \cite{Auer02}.
For a parameter $\alpha$, the \UCB{} indexes are computed as follows, as a sum of
the \emph{average reward} $\widehat{\mu_k}(t)=\frac{X_k(t)}{N_k(t)}$
and a \emph{confidence radius} $\xi_k(t) = \sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}$.
%
\begin{align}\label{eq:UCB_index}
    U_k^{\UCB}(t) = \underbrace{\frac{X_k(t)}{N_k(t)}}_{\widehat{\mu_k}(t)} + \underbrace{\sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}}_{\xi_k(t)}.
\end{align}

This selection rule makes sense for the following reason:
an arm $k$ is chosen at step $t$ because it has a large $U_k^{\UCB}(t) = \UCB_k(t)$, which can happen for two reasons:
$i)$ because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high reward,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{}  respectively represent exploitation and exploration, and summing them up is a natural way to trade off the two.
The parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to chose $\alpha=1/2$ for optimal uniform performance.

To analyze the algorithm $\UCB_1$, we denote again $k^*$ an optimal arm, and $A(t)$ is the arm chosen by the algorithm in round $t$.
According to the algorithm, $\UCB_k(t) \geq \UCB_{k^*}(t)$.
%
Under the clean event,μ(at) +rt(at)≥ ̄μt(at)andUCBt(a∗)≥μ(a∗). Therefore:
XXX
It follows that
XXX

More details on the proofs can be found for instance in Chapter~2 of \cite{Bubeck12}, or Section~1.3 of \cite{Slivkins2019}.
% \TODOL{Expliquer la fin}
The $\UCB_1$ algorithm achieves order-optimal problem-dependent bounds with a $\cO(K\log(T)/\Delta^2)$ regret.

@@

Optimal upper confidence bounds based on Kullback-Leibler divergence and klUCB indexes

If rewards lie in $[a,b]$, then the indexes are computed as follows:
\begin{align}\label{eq:klUCB_index}
    U_k^{\klUCB}(t) = \sup\limits_{q \in [a, b]} \left\{ q : \kl\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{c \log(t)}{N_k(t)} \right\}.
\end{align}

\TODOL{Short note on how to implement such policy, using Newton's method ?}

>>>>

We give an overview of the main families of algorithms designed to tackle stochastic MAB problems, and we mainly focus on the \UCB, \klUCB{} and Thompson sampling algorithms, as they are used in the rest of the manuscript.
% We present SMPyBandits in the next Chapter~\ref{chapter:3}, a Python open-source library, designed to run numerical simulations of MAB problems, one of the main contributions of this thesis.
% We use it to compare the empirical performances of the most famous and most efficient algorithms, in terms of regret and by evaluating them on different problems.
% %
% In the point-of-view of numerical simulations or real-world applications of bandits on embedded systems, one must also take care of two other measures of performance: time and memory complexities, which can vary greatly between different families of algorithms.
% We also empirically compare different algorithms for these two measures in Section~\ref{sec:3:timeAndMemoryCosts}.
%
We then give regret upper-bounds of the two main algorithms used in this thesis, \UCB{} and \klUCB, to highlight the difference between their finite-time regret upper-bounds. They are order-optimal as they attain a logarithmic regret.
% proving their order optimality.
% (\ie, their regret upper-bound asymptotically matches the lower-bound up-to a constant factor).
% Moreover, \klUCB{} is known to be optimal for a wide range of arms distributions.
% A sketch of the proof of both results is given in Appendix.

@@

\paragraph{Decisions can (and should) depend on the past observations.}

@@

% Other popular applications include the following.
%
% \begin{itemize}
%     \item
    \textbf{A/B testing}, for instance for websites, is a popular application of the best arm identification problem,
    where the task is purely exploratory and the player is asked to identify the best of two options (or more), in a finite number of steps \cite{audibert2010best}.
    The problem can either consider a fixed budget and no freedom on the ending of the game (\ie, an known horizon $T$), or a fixed confidence and a certain freedom on the budget (\ie, the identified arm must be the true best arm with probability at least $1-\delta$) \cite{Garivier16BAI}.
    The theoretical complexity of this problem
    % use of bandit for A/B testing
    was first studied in \cite{Kaufmann14},
    and a more practical point-of-view was proposed in \cite{Jamieson17ABTest}.

@@

\begin{lemma}[Regret decomposition]\label{lem:2:RegretDecomposition}
    The (mean) regret $R_T^{\cA}$ can be decomposed as a sum of the number of selections of sub-optimal arms $k$,
    weighted by their gap, like this:
    \begin{equation}\label{eq:2:RegretDecomposition}
        R_T^{\cA} = \sum_{k=1}^K \Delta_k \; \E[ N_k(T) ]
        = \sum_{\substack{k=1,\dots,K\\\Delta_k > 0}} \Delta_k \; \E[ N_k(T) ].
    \end{equation}
\end{lemma}
%
\begin{proof}\label{proof:2:RegretDecomposition}
    We use the chain rule of expectation, and because the expectation is taken on the randomness of the (\iid) rewards $(r_k(t))_t$ and on the decisions of the player $(A(t))_t$ (which are measurable wrt to the past observations $\cO_t$),
    we can rewrite the expected cumulated rewards:
    \begin{align*}
        \E \left[ \sum_{t=1}^T r_{A(t)}(t) \right]
        &= \E \left[ \sum_{k=1}^K \sum_{t=1}^T r_k(t) \mathbbm{1}(A(t) = k) \right] \\
        &= \sum_{k=1}^K \E \left[ r_k(1) \sum_{t=1}^T \mathbbm{1}(A(t) = k) \right]
        % \underbrace{XXX}_{= N_k(T)}
        = \sum_{k=1}^K \mu_k \E \left[ N_k(T) \right].
    \end{align*}
    % FIXME there is a conditionning to do here!
    %
    Thus $R_T^{\cA} = \sum_{k=1}^K \Delta_k \E[ N_k(T) ]$.
    The sum can then be simplified to only account for sub-optimal arms, \ie, arms $k$ such that $\Delta_k > 0$.
\end{proof}

@@

% - Worst case regret (max $R_T^{r}$ for $r$ index of Monte-Carlo simulation?)
The \emph{switching cost} $SC^{\cA}(T)$ counts how many times the player's decision has changed from one time step to the next one, \ie, $SC^{\cA}(T) = \sum_{t=1}^{T-1} \indic(A(t) \neq A(t+1))$.
It has recently gained interest in the literature, for instance in \cite{Koren17}.
% FIXME give the main result proved in this paper?
In single-player models,
it is easy to show that achieving logarithmic regret directly implies a logarithmic upper-bound on the switching cost, and conversely the lower-bound from \cite{LaiRobbins85} also gives an asymptotic logarithmic lower-bound.
But even if $SC^{\cA}(T) = \Theta(\log T)$ for an efficient algorithm, it can be interesting to numerically evaluate this quantity, as a large value might indicate an algorithm that is alternating too much between the optimal arm and other arms.
%
% FIXME we give notations but I don't use them in Chapter5 
It is more interesting to studying $SC_M^{\cA}(T)$, the sum of the switching costs of the $M$ players in a multi-player bandit game, as in this case, it is possible that the players follow an efficient centralized or decentralized algorithm (which gives a logarithmic centralized regret) while still suffering from a linear switching cost.
It is very satisfying to obtain a logarithmic switching cost for our algorithm \MCTopM-\klUCB{} in Chapter~\ref{chapter:5}, even if minimizing the switching cost was not one of our objective, and the bound we obtain on $SC_M^{\cA}(T)$ is just a consequence of our proof of the control we give on the regret $R^{\cA}(T)$.
% - Switching cost? \cite{modiDemo2016} \cite{Koren17}
On a more experimental note, it was showed in \cite{modiDemo2016} that the previous state-of-the-art policy for multi-player bandits, \rhoRand, could be tuned to run in batches\footnote{~The batch bandit setting implies that all players use the same decisions for instance for $50$ consecutive times, and update their decisions only once every $50$ time steps. For more details, see \cite{modiDemo2016} for experiments on the multi-player case, or \cite{perchet2016,gao2019batched,kolnogorov2019multi} for theoretical developments on the single-player case.}, in order to reduce by a certain multiplicative factor its switching cost while only adding an additive factor on its regret.

@@

Let us first describe two naive strategies, that both fail dramatically.
They are detailed in Algorithm~\ref{algo:2:naiveStrategies} below.
We recall the notations introduced above in Section~\ref{par:2:interactiveDemoDiscoverMAB}, the sums of rewards are $X_k(t) = \sum_{s=1}^t r_k(s) \mathbbm{1}(A(s) = k)$, and the numbers of samples are $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$.
%
The estimated means, or empirical averages, are $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$).

-- The uniform strategy always plays the $K$ arms uniformly at random.
The player \textcolor{darkgreen}{\emph{only explores}} without using the collected information, and this strategy fails dramatically for any non trivial problem (\ie, linear regret).
Indeed it obtains a linear (mean) regret $R_T = \frac{1}{K} \sum_{k=1}^K \Delta_k T$
which gives $R_T \propto T$ for any problem with at least one sub-optimal arms (\ie, all problems expect where $\mu_1=\mu_2=\dots=\mu_K$).

% Why Follow-the-Leader (EmpiricalMeans https://smpybandits.github.io/docs/Policies.EmpiricalMeans.html) don't work, example.
-- The ``\emph{Follow-the-Leader}'' strategy consists in first playing once each arm, then always playing $A(t)\in\argmax \widehat{\mu_k}(t)$.
The player \textcolor{blue}{\emph{only exploits}} the collected information, and this strategy can fail dramatically, \ie, obtain linear regret in some problems.
Indeed consider $K=2$ Bernoulli arms of means $\mu_1=1/2$ and $\mu_2=\varepsilon$ where $\varepsilon < 1/2$, then with probability $\varepsilon/2$ the player observes a reward of $0$ for arm $1$ then $1$ for arm $2$ on the first rounds, and so she will play arm $2$ for the $T-2$ remaining rounds, giving a linear (mean) regret $R_T \geq \frac{\varepsilon}{2}\left(\frac{1}{2} - \varepsilon\right) (T-1)$.

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, 2, \dots, T$}{
            \uIf{\textcolor{darkgreen}{pure exploration}}{
                Play uniformly at random \textcolor{darkgreen}{$A(t) \in \cU(\{1,\dots,K\})$}
                \tcp*[f]{Explore}
            }
            \uIf{\textcolor{blue}{pure exploitation}}{
                Play uniformly among the arms of maximal empirical mean: \textcolor{blue}{$A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} \widehat{\mu_k}(t))$}
                \tcp*[f]{Exploit}\\
                Observe a reward $r_{A(t)}(t)$, and update $X_k(t)$, $N_k(t)$ and $\widehat{\mu_k}(t)$
            }
		}
		\caption{Naive strategies: \textcolor{darkgreen}{pure exploitation} or \textcolor{blue}{pure exploration}.}
		\label{algo:2:naiveStrategies}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

As illustrated by the two previous examples, an efficient strategy needs to solve the trade-off between exploration and exploitation.
The two following solutions both consist in splitting the $T$ time steps into $T_0$ steps of exploration and $T-T_0$ steps of exploitation.
%  either in an alternative way or in a fixed way (``explore then exploit'').
They are detailed in Algorithm~\ref{algo:2:naiveStrategies}.

% https://smpybandits.github.io/docs/Policies.EpsilonGreedy.html
-- The \textcolor{deeppurple}{$\varepsilon$-greedy strategy} consists in alternating exploration and exploitation at a certain ratio \cite{SuttonBarto2018,Bubeck12,LattimoreBanditAlgorithmsBook}.
Fix $0<\varepsilon<1$, then at each round, with probability $\varepsilon$ the player selects an arm uniformly at random (exploration) and with probability $1-\varepsilon$ the arm with highest empirical mean is selected (exploitation).
If $\varepsilon$ is constant, then the (mean) regret is still growing linearly as it is lower bounded by $\varepsilon \frac{1}{K} \sum_{k=1}^K \Delta_k T$.
In this first case, in average $T_0 = \varepsilon T$ steps are spent on exploration.
%
But if we consider $\varepsilon_t$ decreasing with time $t$, for instance $\varepsilon_t = \varepsilon_0 / t$ with $\varepsilon_0 = 6 K / d^2$, for a constant $0 < d < \min_{k: \Delta_k > 0} \Delta_k$,
then it was shown in \cite{Auer02} that the regret is of the order of $K \log(T) / d + \smallO{T}$, which leads to an order-optimal regret of $\bigO{\log(T)}$.
In this second case, an average of $T_0 \leq \varepsilon_0 \log(T)$ steps are spent on exploration.

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, 2, \dots, T$}{
            \uIf{\textcolor{deeppurple}{$\varepsilon$-greedy}}{
                Sample a value uniformly in $[0,1]$: $U_t\sim\cU([0,1])$\\
                \uIf{\textcolor{deeppurple}{$U_t < \varepsilon$ (\ie, with probability $\varepsilon$)}}{
                    Play uniformly at random $A(t) \in \cU(\{1,\dots,K\})$
                    \tcp*[f]{Explore}
                }
                \uElse{
                    Play uniformly among the arms of maximal empirical mean: $A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} \widehat{\mu_k}(t))$
                    \tcp*[f]{or Exploit}
                }
            }
            \uIf{\textcolor{deepgold}{Explore-then-Exploit}}{
                \uIf{$t \leq \textcolor{deepgold}{T_0}$}{
                    Play uniformly at random $A(t) \in \cU(\{1,\dots,K\})$
                    \tcp*[f]{Explore}
                }
                \uElse{
                    \uIf{$t = \textcolor{deepgold}{T_0}$}{
                        Pick one arm $A(\textcolor{deepgold}{T_0})=k$ uniformly among the arms of maximal empirical mean: \textcolor{deepgold}{$A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} \widehat{\mu_k}(\textcolor{deepgold}{T_0}))$}
                        \tcp*[f]{then Exploit}
                    }
                    Play the same arm $A(t) = A(\textcolor{deepgold}{T_0})$
                }
            }
            Observe a reward $r_{A(t)}(t)$, and update $X_k(t)$, $N_k(t)$ and $\widehat{\mu_k}(t)$
		}
		\caption{Simple efficient strategies: \textcolor{deeppurple}{$\varepsilon$-greedy} and \textcolor{deepgold}{Explore-then-Exploit}.}
		\label{algo:2:naiveStrategies}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

% https://smpybandits.github.io/docs/Policies.ExploreThenCommit.html
-- The ``\textcolor{deepgold}{Explore-then-Exploit}'' strategy, as studied for instance in \cite{GarivierETC2016}, consists in first exploring uniformly the $K$ arms for $T_0$ time steps, then only exploiting the best arm.
We can lower bound the regret of this strategy by
$R_T \geq K (\min_{k: \Delta_k > 0} \Delta_k) T_0/K + p (T - T_0)$,
if $p$ denotes the probability that the chosen arm is not optimal (\ie, the arm with highest mean after $T_0/K$ samples of each arms).
And we can show that $p$ only depends on the number of collected samples $T_0$ and the gaps $\Delta_k$,
so if $T_0$ is fixed independently of $T$ and the problem difficulty, the regret is again growing linearly, \ie, $R_T = \Omega(T)$.

\begin{proof}
    We prove it for the case of two arms.
    If the distributions are Bernoulli (or sub-Gaussian), we can use Hoeffding's inequality from \cite{hoeffding1963probability} to control this probability $p$.
    %
    Indeed if $Z_1,\dots,Z_n$ are \iid{} samples from a Bernoulli distribution of parameter $\mu$, of mean $\widehat{Z_n} = \frac{1}{n} \sum_{i=1}^n Z_i$, Hoeffding's inequality gives both
    \begin{equation}
        \forall x < \mu, \bP(\widehat{Z_n} < x) \leq \exp(-2 n (x-\mu)^2),
        \;\;\;
        \forall x > \mu, \bP(\widehat{Z_n} > x) \leq \exp(-2 n (x-\mu)^2).
    \end{equation}
    %
    If $\mu_1 = \mu_2 + \Delta$, then
    $p = \bP(\widehat{\mu_1} < \widehat{\mu_2}) \leq \bP(\widehat{\mu_1} < \mu_1 - \Delta/2) + \bP(\widehat{\mu_2} > \mu_2 + \Delta/2)$, and both terms can be bounded by using Hoeffding's inequality with $n=T_0/2$ samples,
    to obtain $p \leq 2 \exp(-T_0 \Delta^2 / 4)$.

    A simple union bound extends the proof for $K$ arms, giving
    $p \leq 2 K \exp(-T_0 d^2 / 4)$, if $d$ is a lower-bound on the positive gaps.
    %
    In all cases, it shows that $p$ is a constant wrt $T$.
\end{proof}

@@

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, 2, \dots, T$}{
            \uIf{$t \leq K$}{
                Play arm $A(t) = t$\;
            }
			\Else(){
                For each arm, compute the \textcolor{blue}{index $U_k(t)$}\;
                Play arm $A(t)$ uniformly among the arms of maximal index: \textcolor{blue}{$A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} U_k(t))$}\;
            }
            Observe a reward $r_{A(t)}(t)$ and update $X_k(t)$, $N_k(t)$ and $\widehat{\mu_k}(t)$\;
		}
		\caption{A generic \textcolor{blue}{index} policy, using \textcolor{blue}{indexes $U_k(t)$} (\eg, \UCB, \klUCB{} etc).}
		\label{algo:2:indexPolicy}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

Let us consider another approach for adaptive exploration, known as ``optimism under uncertainty'': assume each arm is as good as it can possibly be given the observations so far, and choose the best arm based on these optimistic estimate.
%
This intuition leads to the $\UCB_1$ algorithm, initially introduced in \cite{Auer02}.
For a parameter $\alpha$, the \UCB{} indexes are computed as follows, as a sum of
the \emph{average reward} $\widehat{\mu_k}(t)=\frac{X_k(t)}{N_k(t)}$
and a \emph{confidence radius} $\xi_k(t) = \sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}$.
%
\begin{align}\label{eq:UCB_index}
    U_k^{\UCB_1}(t) = \UCB_k(t) = \underbrace{\frac{X_k(t)}{N_k(t)}}_{\widehat{\mu_k}(t)} + \underbrace{\sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}}_{\xi_k(t)}.
\end{align}

This selection rule \eqref{eq:UCB_index} makes sense for the following reason:
an arm $k$ is chosen at step $t$ because it has a large $\UCB_k(t)$, which can happen for two reasons:
$i)$ because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high mean reward $\mu_k$,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
One can also observe that $\xi_k(t)\to\infty$ if $t\to\infty$ but $N_k(t)\ll\log(t)$, thus all arms must be explored at least $\mathcal{O}(\log(T))$ times.
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{}  respectively represent exploitation and exploration, and summing them up is a natural way of trading them off.
The parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to choose $\alpha=1/2$ for optimal uniform performance.

To analyze the algorithm $\UCB_1$, we denote again $k^*$ an optimal arm, and $A(t)$ is the arm chosen by the algorithm in round $t$.
According to the algorithm, $\UCB_k(t) \geq \UCB_{k^*}(t)$.
%
Under the clean event,μ(at) +rt(at)≥ ̄μt(at)andUCBt(a∗)≥μ(a∗). Therefore:
XXX
It follows that
XXX
\TODOL{Expliquer la fin}

More details on the proofs can be found for instance in Chapter~2 of \cite{Bubeck12}, or Section~1.3 of \cite{Slivkins2019}.
The $\UCB_1$ algorithm achieves order-optimal problem-dependent bounds with a $\cO(K\log(T)/\Delta^2)$ regret.

@@

The \klUCB{} algorithm XXX
Optimal upper confidence bounds based on Kullback-Leibler divergence and klUCB indexes

% If rewards lie in $[a,b]$, then t
The indexes are computed as follows, for an exploration function $f(t)$
that is typically chosen as $f(t)=\log(t)+3\log(\log(t))$ for the theoretical analyses,
and $f(t)=\log(t)$ in practice.
The constant $c>0$ is usually chosen as $c=1$.
\begin{align}\label{eq:klUCB_index}
    U_k^{\klUCB}(t) = \sup\limits_{q \in [0, 1]} \left\{ q : \KL\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{c f(t}{N_k(t)} \right\}.
\end{align}
